PHAT RAID, yo

### 2014-05-14 paper/paper.tex
Bunch of changes to the paper

@@

\phatraid{} is implemented in Erlang. Erlang is a pure, functional language with
extensive library support for distributed applications. Erlang has one unit of
concurrency, the process, and one unit of distribution, the node. Each Erlang
node consists of an arbitrary number of processes. Erlang nodes may live on the
same machine or on different machines connected by a network. Every Erlang node
is given a name by an Erlang name server. Erlang processes may communicate with
any Erlang node which they can name. Erlang processes interact with one another
by making RPC calls.

@@

Each \phat{} cluster maintains a file system which is separate and distinct from
the file systems of other \phat{} clusters. Generally, two phat clusters do not
communicate.

@@

A \phatraidcf{$C$}{$f$} implementation is parameterized by two numbers: $C$, the
number of \phat{} clusters, and $f$, the \phat{} cluster failure threshold. A
single \phat{} cluster contains $2f + 1$ nodes. A \phatraid{} group contains $C$
clusters for a total of $C(2f+1)$ \phat{} nodes. A \phatraid{} group can be
concurrently accessed by an arbitrary number of raid-clients.

On a client node, the raid-client spawns $C$ client processes which each maintain
a connection to one of the $C$ \phat{} clusters. All communications between the
raid-client and a phat cluster go through the assigned client process.

@@

In Erlang, the basic unit of concurrency is a \emph{process} and an Erlang VM is
known as a \emph{node}. A process may communicate with other processes on its
node as well as processes on other nodes. Multiple nodes may be on the same
machine or distributed across many machines.

The server cluster is implemented as a collection of nodes running the same
server code. One of these nodes is designated the master node. The other nodes
are known as replicas.

The master and each replica has four components: the supervisor, the
(unfortunately named) server, VR, and the file system.

\begin{itemize}
\item the supervisor -- if any of the other three processes dies, it shuts all
  of them down and restarts all of them
\item the server -- if the node is the master, it passes messages on to VR; if
  the node is a replica, it redirects clients to the master
\item VR -- maintains a log of opaque messages, achieving consensus with other
  nodes in the cluster via the Viewstamped Replication protocol
\item the file system -- a simple hierarchical file system with locks
\end{itemize}

@@

\begin{itemize}
\item The \texttt{supervisor} behavior provides a framework for restarting
  failed processes.
\item The \texttt{gen\_server} behavior provides a framework for many-client
  single-server interactions. We implemented functions to process messages
  formatted as Erlang tuples. The \texttt{gen\_server} behavior handles
  low-level socket interaction.
\item The \texttt{gen\_fsm} behavior generalizes the \texttt{gen\_server}
  behavior by permitting the server to have a finite number of states. Each
  state has a set of functions with which to process messages. This behavior can
  be seen as an ad hoc polymorphic variant of the \texttt{gen\_server} behavior.
\end{itemize}

@@

Each child in the list of children is a 6-tuple consisting of: child name,
initial procedure call, child transience, shutdown style, child type, necessary
modules.

\begin{itemize}
\item The initial procedure call is a 3-tuple of module, procedure name and
  argument list.
\item Every child's transience is specified as \lstinline!permanent!, meaning
  they should be restarted.
\item Every child has shutdown style \lstinline!1! meaning they will first be
  asked to terminate and one second later they will be forcibly terminated.
\item All of our children are workers, which are distinguished from
  sub-supervisors.
\item The necessary modules are dynamically loaded when the child processes is
  initialized, there is a one-to-one mapping between our children and code
  modules.
\end{itemize}

@@

We implemented Viewstamped Replication\cite{liskov2012viewstamped} to achieve
consensus in the file system. Listing \ref{lst:genfsm} defines how to handle
prepare messages sent to a node in the replica state. It is piecewise defined by
pattern matching on the arguments. The first two definition clauses use the
\lstinline!when! keyword to specify arbitrary required relationships between the
matched variables.

The final definition clause triggers when the replica receives a message in the
current view. The replica adds the message to a sorted queue and calls a
processing procedure. The processing procedure is defined separately in Listing
\ref{lst:processing1} and Listing \ref{lst:processing2}.

Listing \ref{lst:processing1} handles a message from an out-of-date master,
i.e., his commit number is older than the replica's commit number. The replica
first proposes a view change to the cluster. Afterwards, it changes its
\texttt{gen\_fsm} state to \lstinline!viewChange! and updates the
store\footnote{The store is called \lstinline!State! in the code listings} with
the new view number and the new master.

Listing \ref{lst:processing2} handles messages from the current master. The
\lstinline!prepareBuffer! stores prepare messages that arrived out of order. In
particular, if the log ends at operation number $n$ and message $n+1$ is dropped
by the network, all subsequent messages will be buffered. The buffered messages
will not be added to the log until message $n+1$ is received. The call to
\lstinline!processBuffer! moves messages to the log, if all previous messages
have now been received. The call to \lstinline!processLog! commits new messages
in the log which have been committed by the master node. Lines 405 to 408
destructure \lstinline!AfterLog! and bind variables for later use. The final
\lstinline!if! statement responds to the master unless the master has committed
messages we haven't received.\footnote{This can happen if the network drops a message whose
  operation number lies between the replica's last commit number and the
  master's last commit number.}

@@

We distinguish between two types of failure: cessation of progress and loss of
data. In a VR cluster without persistent stores, data loss only occurs when all
$2f+1$ nodes in the cluster fail. In a VR cluster, progress only ceases if $f+1$
nodes are lost. In \phatraid{} the conditions for cessation of progress and loss
of data are complicated by the combination of RAID and VR.

@@

Four distributions of failed nodes in \phatraidcf{3}{1} are depicted in Figure
\ref{fig:failures}. In this section, if we claim ``data is lost'', we mean that
if no persistent store is used, data will be lost. Adding a persistent store
obviously prevents loss of committed data.

The top left corner depicts five of the nine nodes failing. This state is
\emph{not} a failure state. The group can still make progress and no data is
lost. The two surviving clusters can reach consensus on new values and can
reconstruct the failed cluster's data.

The top right corner depicts four of the nine nodes failing. This state is a
failure state. The group cannot make progress but no data can been
lost. Progress is blocked because more than one cluster cannot reach
consensus. Since no cluster was completely destroyed, all data still exists on
at least one node.

The bottom left corner is impossible because if data can be lost then we
certainly cannot make progress.

The bottom right corner depicts six of the nine nodes failing. This state is
also a failure state. The group cannot make progress and will lose
data. Progress is blocked because more than one cluster cannot reach
consensus. Additionally, two complete clusters have been lost so half of the
data cannot be recovered.

In general, progress ceases if more than one cluster cannot reach consensus and
data loss occurs if more than one cluster completely fails. The lower bound of
failed nodes for progress cessation in \phatraidcf{$C$}{$f$} is:
    $$ 2\cdot(f+1) $$
and the lower bound of failed nodes for data loss is:
    $$ 2\cdot(2f+1) $$

In contrast, if all data has been propagated to all nodes, data is only lost
when all nodes die. If data has not been propagated to all nodes, then loss
occurs if all nodes containing the data die.

@@

A single number cannot describe the failure threshold of a \phatraid{} cluster
because of the interplay between VR and RAID. The failure of the \phatraid{}
group is dependent on the arrangement of the failed nodes. For example, Figure
\ref{fig:failures_noprogress} depicts a situation wherein only four failures
stops progress. In contrast, Figure \ref{fig:failures_allok} depicts a situation
wherein five failures does \emph{not} stop progress.

In a \phatraidcf{3,1} group, less than one-quarter of four-failure arrangements
stop progress. The necessary condition for progress cessation is that two or
more clusters have lost at least $f+1$ nodes. By the pidgeon hole principle,
only the two-in-two, four-failure arrangement stops progress. We can count the
number of such arrangements:

@@

When five nodes fail, they can be distributed into each group in the
proportions: (3,1,1), (3,2,0) and (2,2,1). The first case does not stop progress
because only one cluster fails. The later two cases stop progress because two
clusters fail. We calculate the probability of \phatraid{} group failure when
five nodes fail:

\begin{align*}
  \textrm{(3,1,1)}\;\;\;
    &\binom{3}{1}\binom{3}{3}\binom{3}{1}\binom{3}{1} = 3^3 = 27 \\
  \textrm{(3,2,0)}\;\;\;
    &\binom{3}{1}\binom{2}{1}\binom{3}{3}\binom{3}{2} = 2\cdot3^2 = 18 \\
  \textrm{(2,2,1)}\;\;\;
    &\binom{3}{1}\binom{3}{2}\binom{3}{1}\binom{3}{1} = 3^4 = 81 \\
  \textrm{5-Failure Arrs.}\;\;\;
    &\binom{9}{5} = 126 \\
  \textrm{Failure Prob.}\;\;\;
    &\Pr\left[\textrm{group failure} \,|\, n_{failed} = 5\right] = \frac{81 + 18}{126} =
    \frac{99}{126} \approx 0.79 \\
\end{align*}

@@

The general formula for the failure probability distribution is not clear, but
it is clear that a \phatraidcf{$C$}{$f$} group is absolutely resilient to $2(f +
1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
to $2(f+1)$ failures and unlikely resistant to $2(f+1)+1$ failures.

The general form for the probability of \phatraidcf{$C$}{$f$} group failure
given $2(f + 1)$ node failures is:

\begin{align*}
  \Pr\left[\textrm{group failure} \,|\, \phatraidcf{C}{f}, n_{failure}=2(f+1)\right]
    =
  \frac{\binom{C}{2}\binom{2f+1}{f+1}^2}
       {\binom{C \cdot (2f+1)}{2(f+1)}}
\end{align*}

@@

\begin{align*}
  P\left[\textrm{group failure} \,|\, \phatraidcf{6}{1}, n_{failure}=4\right]
    = \frac{135}{3000} \approx 0.05
\end{align*}

@@

Extra parity bits!

@@

>>>>

Google's \chubby{}\cite{burrows-chubby} is a distributed file system which
achieves consensus in the face of limited failures. Unfortunately, for storing
large files disk latency on the cluster nodes is a bottleneck. We propose
building a RAID group out of distributed file system clusters. Large files are
sharded and disk latency is reduced proportional to number of clusters. Failure
resiliency is not completely sacrificed and we provide an analysis of the
new failure modes.

@@

\phatraid{} is implemented in Erlang. Erlang is a pure, functional language with
extensive library support for distributed applications. Erlang has one unit of
concurrency, the \emph{process}, and one unit of distribution, the
\emph{node}. Each Erlang node consists of an arbitrary number of
processes. Erlang nodes may live on the same machine or on different machines
connected by a network. Every Erlang node is given a name by an Erlang name
server. Erlang processes may communicate with any Erlang node or process it can
name. Erlang processes interact with one another by making RPC calls or sending
messages. Identifiers beginning with a lowercase letter are actually atoms, like
Scheme's symbols. Identifiers beginning with an uppercase letter are normal
identifiers.

@@

Each \phat{} cluster maintains a file system which is separate and distinct from
the file systems of other \phat{} clusters. Generally, two phat clusters do not
communicate. The raid-client and client processes all live on one node. Each
\phat{} master and \phat{} replica lives on its own node.

@@

A \phatraidcf{$C$}{$f$} implementation is parameterized by two numbers: $C$, the
number of \phat{} clusters, and $f$, the \phat{} cluster failure tolerance. A
single \phat{} cluster contains $2f + 1$ nodes. A \phatraid{} group contains $C$
clusters for a total of $C(2f+1)$ \phat{} nodes. A \phatraid{} group can be
concurrently accessed by an arbitrary number of raid-clients.

On a client node, the raid-client spawns $C$ client processes which each
maintain a connection to one of the $C$ \phat{} clusters. Each client process
mediates all communication between its assigned \phat{} cluster and the
raid-client.

A store operation on the raid-client breaks the file into $C-1$ chunks, stores
each chunk on a cluster, and stores a parity bit on the remaining cluster. A
raid-client fetch operation reconstructs the file from the $C-1$ chunks, using
the parity bit if necessary.

@@

The server cluster is a collection of nodes running the same server code. One of
these nodes is designated the master node (aka \phat{} Master). The other nodes are
known as \phat{} replicas or just replicas.

The master and each replica has four components: the supervisor, the
(unfortunately named) server, VR, and the file system:

\begin{itemize}
\item the supervisor -- restarts the other three processes if any single node dies
\item the server -- passes messages to the VR process if this node is the
  \phat{} master; otherwise, it drops the message and informs the client of the
  current master node
\item VR -- maintains a log of opaque messages, achieving consensus with other
  nodes in the cluster via the Viewstamped Replication protocol
\item the file system -- maintains a simple hierarchical file system with locks
\end{itemize}

@@

\begin{itemize}
\item The \texttt{supervisor} behavior provides a framework for restarting
  failed processes.
\item The \texttt{gen\_server} behavior provides a framework for many-client
  single-server interactions. We implemented functions to process messages
  formatted as Erlang tuples. The \texttt{gen\_server} behavior handles
  low-level socket interaction, messaging queueing, etc.
\item The \texttt{gen\_fsm} behavior generalizes the \texttt{gen\_server}
  behavior by permitting the server to have a finite number of states. Each
  state has a set of functions with which to process messages. This behavior can
  be seen as an ad hoc polymorphic variant of the \texttt{gen\_server} behavior.
\end{itemize}

@@

Each child in the list of children is a 6-tuple consisting of: child name,
initial procedure call, child transience, shutdown style, child type, and
necessary modules.

\begin{itemize}
\item The initial procedure call is a 3-tuple of module, procedure name and
  argument list.
\item Every child's transience is specified as \lstinline!permanent!, meaning
  they \emph{should} be restarted.
\item Every child has shutdown style \lstinline!1! meaning they will first be
  asked to terminate and one second later they will be forcibly
  terminated\footnote{We could also have specified the hastier shutdown style:
    \lstinline!brutal_kill!}.
\item All of our children are workers, which are distinguished from
  sub-supervisors.
\item The necessary modules are dynamically loaded when the child processes is
  initialized. In \phatraid{}, each child only depends on one module.
\end{itemize}

@@

The \phat{} file system achieves consensus via the Viewstamped Replication
protocol \cite{liskov2012viewstamped}. Listing \ref{lst:genfsm} defines how to
handle prepare messages sent to a node in the replica state. It is piecewise
defined by pattern matching on the arguments. The first two definition clauses
use the \lstinline!when! keyword to specify arbitrary required relationships
between the matched variables.

The final definition clause triggers when the replica receives a message in the
current view. The replica adds the message to a sorted queue and calls a
processing procedure. The processing procedure is defined separately in Listings
\ref{lst:processing1} and \ref{lst:processing2}.

Listing \ref{lst:processing1} handles a message from an out-of-date master,
i.e., the master's commit number is older than the replica's commit number. The
replica first proposes a view change to the cluster. Afterwards, it changes its
\texttt{gen\_fsm} state to \lstinline!viewChange! and updates the
store\footnote{The store is called \lstinline!State! in the code listings} with
the new view number and the new master.

Listing \ref{lst:processing2} handles messages from the current master. The
\lstinline!prepareBuffer! stores prepare messages that arrived out of order. In
particular, if the log ends at operation number $n$ and message $n+1$ is dropped
by the network, all subsequent messages will be buffered. The buffered messages
will not be added to the log until message $n+1$ is received. The call to
\lstinline!processBuffer! moves messages to the log, if all previous messages
have now been received. The call to \lstinline!processLog! commits logged
messages which now have been committed by the master node. Lines 405 to 408
destructure \lstinline!AfterLog! and bind variables for later use. The final
\lstinline!if! statement responds to the master unless the master has committed
messages we haven't received.\footnote{This can happen if the network drops a
  message whose operation number lies between the replica's last commit number
  and the master's last commit number.}

@@

We distinguish between two types of failure: cessation of progress and loss of
data. In a VR cluster without persistent stores, data loss of fully propagated,
commited data only occurs when all $2f+1$ nodes in the cluster fail \footnote{If
  more than $f+1$ nodes die some ``commited but not fully propogated data'' can
  be lost}. In a VR cluster, progress only ceases if $f+1$ nodes are lost. In
\phatraid{} the conditions for cessation of progress and loss of data are
complicated by the combination of RAID and VR.

@@

Four arrangements of failed nodes in \phatraidcf{3}{1} are depicted in Figure
\ref{fig:failures}. In this section, if we claim ``data is lost'', we mean that
if no persistent store is used, data will be lost. Adding a persistent store
obviously prevents loss of committed data. Additionally, we assume node failures
are independent and uniformly distributed throughout the \phatraid{} group.

The top left corner, Figure \ref{fig:failures_allok}, depicts five of the nine
nodes failing. This state is \emph{not} a failure state. The group can still
make progress and no data is lost. The two surviving clusters can reach
consensus on new values and can reconstruct the failed cluster's data.

The top right corner, Figure \ref{fig:failures_noprogress}, depicts four of the
nine nodes failing. This state is a failure state. The group cannot make
progress; however, no data can been lost. Progress is blocked because more than
one cluster cannot reach consensus. Since no cluster was completely destroyed,
all data still exists on at least one node.

The bottom left corner, Figure \ref{fig:failures_impossible}, is impossible
because if data can be lost then we certainly cannot make progress.

The bottom right corner, Figure \ref{fig:failures_noprogress_dataloss}, depicts
six of the nine nodes failing. This state is also a failure state. The group
cannot make progress and will lose data. Progress is blocked because more than
one cluster cannot reach consensus. Additionally, two complete clusters have
been lost so half of the data cannot be recovered.

@@

In general, progress ceases if more than one cluster \emph{cannot reach
  consensus} and data loss occurs if more than one \emph{cluster completely
  fails}. The lower bound of failed nodes for progress cessation in
\phatraidcf{$C$}{$f$} is:

    $$ 2\cdot(f+1) $$

and the lower bound of failed nodes for data loss is:

    $$ 2\cdot(2f+1) $$

By the pidgeon hole principle, the upper bound of failed nodes which still
permit progress is,

    $$ C\cdot f + 1 $$

because we can place $f$ failures in all but one cluster and place $f+1$
failures in the final cluster.

The upper bound for no data loss is,

    $$ C\cdot 2f + 1 $$

because we can keep one node of all but one cluster alive. From these $C-1$
nodes we can reconstruct the entire file system.

The bounds presented above cannot fully describe the failure threshold of a
\phatraid{} cluster because of the interplay between VR and RAID. Instead, we
will discuss the probability of group failure given $n$ node failures.

Group failure is dependent on the arrangement of failed nodes. For example,
Figure \ref{fig:failures_noprogress} depicts a situation wherein only four
failures stops progress. In contrast, Figure \ref{fig:failures_allok} depicts a
situation wherein five failures does \emph{not} stop progress.

In a \phatraidcf{3}{1} group, less than one-quarter of four-failure arrangements
stop progress. The necessary condition for progress cessation is that two or
more clusters have lost at least $f+1$ nodes. By the pidgeon hole principle,
only the two-in-two, four-failure arrangement stops progress. We can count the
number of such arrangements:

@@

When five nodes fail, they can be distributed into each group: (3,1,1), (3,2,0)
and (2,2,1). The first case does not stop progress because only one cluster
fails. The later two cases stop progress because two clusters fail. We calculate
the probability of \phatraid{} group failure when five nodes fail:

\begin{align*}
  \textrm{(3,1,1)}\;\;\;
    &\binom{3}{1}\binom{3}{3}\binom{3}{1}\binom{3}{1} = 3^3 = 27 \\
  \textrm{(3,2,0)}\;\;\;
    &\binom{3}{1}\binom{2}{1}\binom{3}{3}\binom{3}{2} = 2\cdot3^2 = 18 \\
  \textrm{(2,2,1)}\;\;\;
    &\binom{3}{1}\binom{3}{2}\binom{3}{1}\binom{3}{1} = 3^4 = 81 \\
  \textrm{5-Failure Arrs.}\;\;\;
    &\binom{9}{5} = 126 \\
  \textrm{Failure Prob.}\;\;\;
    &\Pr\left[\textrm{group failure} \,|\, \phatraidcf{3}{1},\, n_{failed} = 5\right] \\
    &= \frac{81 + 18}{126} = \frac{99}{126} \approx 0.79 \\
\end{align*}

@@

The general formula for the failure probability distribution is not clear, but
it is clear that a \phatraidcf{$C$}{$f$} group is absolutely resilient to $2(f +
1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
to $2(f+1)$ failures and unlikely resistant to $C\cdot f + 1$ failures.

For exactly $2(f+1)$ node failures, the probability of \phatraidcf{$C$}{$f$}
group failure is:

\begin{align*}
  \Pr\left[\textrm{group failure} \,|\, \phatraidcf{C}{f},\, n_{failure}=2(f+1)\right]
    =
  \frac{\binom{C}{2}\binom{2f+1}{f+1}^2}
       {\binom{C \cdot (2f+1)}{2(f+1)}}
\end{align*}

@@

\begin{align*}
  P\left[\textrm{group failure} \,|\, \phatraidcf{6}{1},\, n_{failure}=4\right]
    = \frac{135}{3000} \approx 0.05
\end{align*}

@@

We would like to consider adding more than one parity bit. For some
applications, the diminishing returns on latency are not worth the loss of
failures. In that case, we would do well to add more parity bits, thus allowing
more clusters to completely fail.

@@

We've presented RAID on a distributed file system which shows that sharding
files improves latency for large files while maintaining reasonable failure
probabilities.
